Leadership across cultures and contexts: an exploratory analysis

### 2020-05-10 Response to reviewers.Rmd

This is a good suggestion for future research, but goes beyond remit of the current paper.

**Finally, the reviewers made a number of valuable suggestions, not enumerated above,  that should help you sharpen you presentation and discussion. Please attend to those.**

## Reviewer 1 comments

**This manuscript presents a descriptive analysis of the HRAF to provide information about leadership in traditional societies. Overall, I enjoyed reading the paper and thought it was a useful resource. It provides a comprehensive analysis of leadership across a wide swath of traditional cultures. It reports on the types of functions leaders serve, the qualities possessed by leaders, and the costs and benefits of leadership experienced by both leaders and followers. The paper will likely be of interest to anyone working in the area of leadership, social hierarchy, or group behavior. I provide some comments below aimed at helping the authors increase the potential accessibility and impact of the work.**

@@

We added additional details in the methods section on the eHRAF and ethnographic sample and now more clearly direct readers to the original publication from which the database was built, which includes additional details.

We also include additional details in the results and supplementary information on the corpus of ethnographic documents which provide the sample of ethnographic paragraphs. NEED TO DO.

**I don't work with ethnographic data, so perhaps this question is naïve, but: What conclusions can we draw about the actual frequency of leadership functions and traits from the frequency with which they are mentioned in the ethnographic record? There is an implicit assumption that the two are closely aligned. Has such an alignment been confirmed? Is that even testable?**

We now are careful to frame our results as "evidence for" or "lack of evidence for" leadership traits in each culture, and to make clear that absence of evidence is not evidence of absence.

@@

**The abstract could be tightened up in terms of description of particular costs and benefits to leaders and followers. And a brief justification in the abstract for why the study compares costs and benefits would be useful- e.g. understanding evolution of leader or follower strategies requires understanding the (fitness-relevant) costs and benefits to them.**

Good point. [Need to do]

**Lines 88-89 suggest that the paper's ethnographic analysis will include leadership at all levels of social organization. But does the theory and analysis reported in the paper really extend to all levels of social organization- such as child play groups or households? It seems the focus is still community-level cooperation among adults. Perhaps clarify.**

This statement in the introduction is intended to apply across disciplines focused on leadership dynamics and not specifically apply to the methods of the current study. The reviewer brings up a well-received point, however. Any work in developmental psychology on social hierarchy and leadership would not be subject to this critique. We have not rephrased this claim to better fit with our objectives and outcomes here.

### 2020-05-10 leadership_across_cultures_contexts.Rmd
Benefit cost ratio, mostly

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(feature_models_aic)
loadd(features_support_cult)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_features_support)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

```{r varsupportcb, echo=FALSE, warning=FALSE, message=F, fig.width=15, fig.height=11, fig.cap="Support for each coded variable for the costs and benefits of leadership and followership. Blue: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Red: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablecostsbenefits.pdf", plot_costs_benefits)
plot_costs_benefits
```

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social material success" feature, which included Polygyny, High quality spouse, and Many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for Social material success was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

Four leadership dimensions varied by subsistence strategy. There was more evidence that hunter-gatherer leaders lacked coercive authority compared to other leaders, which supports much theory and ethnography discussing egalitarianism and the active resistance of hierarchy among hunter-gatherers [e.g., @boehm_hierarchy_1999;@boehm_purposive_2008; @woodburn_egalitarian_1982]. Reliance on a mobile and stochastic resource base is suggested to promote strong social norms of sharing [@peterson_demand_1993; @cashdan_egalitarianism_1980] and shape both resource transfers and partner preferences [c.f., @smith_friend_2018; @smith_hunter-gatherers_2018]. There was more evidence that both hunter-gatherer and pastoralist leaders generated food surpluses, but less evidence that pastoralist leaders distributed food and other resources. Thus, systematic ethnographic evidence supports the important role of provisioning resources by high status males among foragers [@wood_prestige_2006; @gurven_give_2005; @gurven_men_2009; @alger_paternal_2020]. However, there was more evidence that leaders in communities with mixed subsistence strategies distributed resources, which lends support to high status males using economic success and producitivity as a status-seeking strategy [@hawkes_showing_2002; @hawkes_family_2010].

There was also evidence of variation between our factors, i.e., the distribution of text records by group context varied by subsistence strategy (see Figure \@ref(fig:grouptype)). Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Evidence for several theoretically important leadership dimensions was relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Mediate, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/material success. Prosocial competencies was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).

>>>>

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(feature_models_aic)
loadd(features_support_cult)
loadd(follower_bc_ratio)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_bc_ratio)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_features_support)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

```{r varsupportcb, fig.width=15, fig.height=11, fig.cap="Support for each coded variable for the costs and benefits of leadership and followership. Blue: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Red: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
ggsave(file="Figures/variablecostsbenefits.pdf", plot_costs_benefits)
plot_costs_benefits
```

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 3)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 3)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

Within benefit/cost dimensions, there was relatively more evidence for mating and territory benefits than costs for both leaders and followers, and equal levels of evidence for risk of harm. For leaders compared to followers, there were noticably higher odds ratios for inclusive fitness benefits, material resources, and social services. See Figure \@ref(fig:benefitcostratio).

```{r benefitcostratio, fig.width=10, fig.cap="The odds ratios of finding evidence of a benefit vs. a cost in a text record for leaders and followers, by each dimension of benefits and costs. Odds ratios estimated using a mixed effects logistic regression model of evidence for (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. Separate models were fit for leaders and followers. X-axis is on a log scale."}
plot_bc_OR
```

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social material success" feature, which included Polygyny, High quality spouse, and Many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for Social, reproductive, and material success was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

movement_hg_pct <- signif(100*sum(all_data$functions_NewSettlement[all_data$subsistence=='hunter gatherers'])/sum(all_data$functions_NewSettlement), 3)

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records.

These factors were confounded, i.e., group context was associated with subsistence strategy, which was associated with continental region. See Figures \@ref(fig:grouptype) and \@ref(fig:regionsubsistence). Specifically, evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Because group context, subsistence strategy, and continental region were confounded (and there was some evidence that leader sex was too), we fit multiple regression models with all four factors included. Interestingly, the `r vary_n`  systematically varying dimensions of leadership tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1).

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

Four leadership dimensions varied by subsistence strategy. There was more evidence that hunter-gatherer leaders lacked coercive authority compared to other leaders, which supports much theory and ethnography discussing egalitarianism and the active resistance of hierarchy among hunter-gatherers [e.g., @boehm_hierarchy_1999;@boehm_purposive_2008; @woodburn_egalitarian_1982]. Reliance on a mobile and stochastic resource base is suggested to promote strong social norms of sharing [@peterson_demand_1993; @cashdan_egalitarianism_1980] and shape both resource transfers and partner preferences [c.f., @smith_friend_2018; @smith_hunter-gatherers_2018]. There was more evidence that both hunter-gatherer and pastoralist leaders generated food surpluses, but less evidence that pastoralist leaders distributed food and other resources. Thus, systematic ethnographic evidence supports the important role of provisioning resources by influential males among foragers [@wood_prestige_2006; @gurven_give_2005; @gurven_men_2009; @alger_paternal_2020]. However, there was more evidence that leaders in communities with mixed subsistence strategies distributed resources, which lends support to males using economic success and producitivity as an influence-seeking strategy [@hawkes_showing_2002; @hawkes_family_2010].

Six leadership dimensions varied by continental region (see Figure \@ref(fig:heatmapregion)). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.

Evidence for several theoretically important leadership dimensions was relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (`r movement_hg_pct`%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained [e.g., @gavrilets_solution_2014; @hooper_theory_2010; @price_evolution_2014; @bastardoz_nature_2018; @hagen_leadership_2019; @garfield_review_2019].

We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading and following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported (see Figure \@ref(fig:varsupportcb)).

The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs. The relative abundance of evidence for leader benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 3)`$), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. See Figure \@ref(fig:benefitcostratio).

@@

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Mediate, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/reproductive/material success. Prosocial competencies was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).
